\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-09-02 swmpr_manu.Rnw
most of the formatting is done, ready to butcher text

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.00\\textwidth')
options(replace.assign=TRUE,width=70,digits=1)
knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
@

\title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
\author{by Marcus W Beck}

@@

\abstract{
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making. One example in the United States is the System-Wide Monitoring Program (SWMP) that was implemented in 1995 by the federally-funded National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. SWMP data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with SWMP environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
}

@@

The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.  These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.  Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \cite{NDBC15}.  Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.  However, the increasing quantity of available information to address relevant questions has contributed to the growth of `big data' where analyses are limited by compututational requirements and identifying the signal from the noise rather than the availability of information. A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \cite{Campbell13,Millie13}.

An invaluable source of monitoring data for coastal environments in the United States is provided by the National Estuarine Research Reserve System (NERRS, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the System-Wide Monitoring Program (SWMP) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The SWMP network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current SWMP data.  NERRS researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \cite{SWMP14}.

This article describes a software package that was developed to address research needs of the NERRS program using the open-source statistical programming language R \cite{RDCT14}.  SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System-Wide monitoring program.  Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.  Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.  The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large datasets.

@@

Four core data elements are collected through the SWMP monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant QAQC flags with the appropriate qualifier.

The CDMO web portal was established to support priority areas of SWMP that focus on the continuation and advancement of data management.  As such, CDMO provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic QAQC procedures.    Prior to any data request to the CDMO, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the SWMP network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the CDMO website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with CDMO before using the data retrieval functions in SWMPr.  Web services are provided by CDMO for direct access to SWMP data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on CDMO web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the SWMP sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

\section{Structure of the SWMPr package}

@@

The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be downloaded from CRAN by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

@@

The SWMPr package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes the package structure, beginning with the retrieval functions, a description of the \texttt{swmpr} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

Two basic approaches using SWMPr are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using CDMO web services. To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.

% table for retrieval functions
<<results = 'asis', echo = FALSE>>=
funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
funcs <- paste0('\\texttt{', funcs, '}')

descrips <- c(
  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.',
  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

latex(
  to_tab[, 'Description', drop = F],
  file = '',
  caption = "Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",
  rowlabel = 'Function',
  colheads = 'Description',
  rowname = to_tab$Functions,
  caption.loc = 'top',
  col.just=c("p{3.5in}"), 
  label = 'tab:retrieve', 
  table.env = FALSE, 
  where = '!tbp'
  )
@

@@

Data retrieval functions to import data directly into R from the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.

@@

The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from CDMO.  This approach is most appropriate for large, specific data requests. The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

@@

\subsection{The \texttt{swmpr} object class}

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).  Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.  The following shows an example of the raw data imported using \texttt{all\_params}.

@@

The attributes for a \texttt{swmpr} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).  These act as metadata that are used internally by many of the package functions and are updated as the data are processed. The attributes are not visible with the raw data but can be viewed as follows.

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class.',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class.',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
  'Station identifier used by NERRS as a string with 7 or 8 characters.',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
  'Indicates if QAQC columns are present in the raw data.',
  'Start and end dates for the data.',
'Timezone of the station using the city/country format\\textsuperscript{a}.',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.'
)

to_tab <- data.frame(Attributes = atts, Class = atts_class, Description = atts_desc, stringsAsFactors = F)
foot <- c('\\textsuperscript{a}\\footnotesize Time zones that do not observe daylight savings are used for \\texttt{swmpr} objects and may not be cities in the United States.  For example, ``America/Jamaica" is used for Eastern Standard Time.')

@@

latex(
  to_tab[, -1, drop = F],
  file = '',
  caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",
  rowlabel = 'Attributes',
  colheads = c('Class', 'Description'),
  rowname = to_tab$Attributes,
  caption.loc = 'top',
  insert.bottom = foot,
  col.just=c("p{0.75in}", "p{3.25in}"), 
  label = 'tab:attributes', 
  table.env = FALSE, 
  where = '!tbp'
  )
@

The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.  A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods available in R also apply to \texttt{swmpr} objects.  Available methods for the \texttt{swmpr} class are described below and can also be viewed:

@@

A sample dataset can be downloaded for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

funcs <- c('comb', 'qaqc', 'qaqcchk', 'rem\\_reps', 'setstep', 'subset')
funcs <- paste0('\\texttt{', funcs, '}')

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove QAQC columns and remove data based on QAQC flag values for a \\texttt{swmpr} object.  Only applies if QAQC columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \\texttt{comb}.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.

@@

funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'map\\_reserve', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
funcs <- paste0('\\texttt{', funcs, '}')

>>>>

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.8\\textwidth')
options(replace.assign=TRUE,width=70,digits=1)
knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
@

\title{\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
\author{by Marcus W Beck}

@@

\abstract{
The System-Wide Monitoring Program (SWMP) was implemented in 1995 by the US National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. Simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and provides an effective approach to link quantitative information with analysis tools that will inform coastal protection and restoration.
}

@@

The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \citep{Glasgow04,Fries08}.  These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.  Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \citep{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \citep{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \citep{NDBC15}.  Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \citep{Glasgow04}.  However, the increasing quantity of available information to address relevant questions has contributed to the growth of `big data' where analyses are limited by compututational requirements and identifying the signal from the noise rather than the availability of information. A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \citep{Campbell13,Millie13}.

An invaluable source of monitoring data for coastal environments in the United States is provided by the National Estuarine Research Reserve System (NERRS, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the System-Wide Monitoring Program (SWMP) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The SWMP network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \citet{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \citet{Caffrey03,Caffrey04}, tidal characteristics \citet{Sanger02}, dissolved oxygen \citet{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current SWMP data.  NERRS researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \citep{SWMP14}.

This article describes a software package that was developed to address research needs of the NERRS program using the open-source statistical programming language R.  \CRANpkg{SWMPr} (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System-Wide monitoring program.  Functions provided by \pkg{SWMPr} address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.  Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.  The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large datasets.

@@

Four core data elements are collected through the SWMP monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the \pkg{SWMPr} package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant QAQC flags with the appropriate qualifier.

The CDMO web portal was established to support priority areas of SWMP that focus on the continuation and advancement of data management.  As such, CDMO provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic QAQC procedures.    Prior to any data request to the CDMO, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the SWMP network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the CDMO website. Alternatively, the \code{site\_codes} (all sites) or \code{site\_codes\_ind} (single site) functions provided by the \pkg{SWMPr} package can be used to view the same information.  As noted below, the computer's IP address must be registered with CDMO before using the data retrieval functions in \pkg{SWMPr}.  Web services are provided by CDMO for direct access to SWMP data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in \pkg{SWMPr} are simple calls to the existing retrieval functions on CDMO web services.  For example, the \code{site\_codes} function in \pkg{SWMPr} uses the \code{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the SWMP sites.  The text below describes the data retrieval functions in more detail, including all other functions available in \pkg{SWMPr}.

\section{Structure of the \pkg{SWMPr} package}

@@

The \pkg{SWMPr} package was developed for use with the R ($\geq$ v3.0.0) statistical programming language. The \pkg{SWMPr} package can be downloaded from CRAN by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \code{library} command.

@@

The \pkg{SWMPr} package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.  Functions for retrieving are used to import the data into R as a \code{"swmpr"} object class.  Functions for organizing and analyzing the data provide methods for working with a \code{"swmpr"} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes the package structure, beginning with the retrieval functions, a description of the \code{"swmpr"} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

Two basic approaches using \pkg{SWMPr} are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using CDMO web services. To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.  The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.

% table for retrieval functions
<<results = 'asis', echo = FALSE>>=
funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
funcs <- paste0('\\code{', funcs, '}')

descrips <- c(
  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\code{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\code{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.',
  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\code{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\code{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\code{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

latex(
  to_tab[, 'Description', drop = F],
  file = '',
  caption = "Retrieval functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",
  rowlabel = 'Function',
  colheads = 'Description',
  rowname = to_tab$Functions,
  caption.loc = 'top',
  col.just=c("p{3.5in}"), 
  label = 'tab:retrieve', 
  table.env = FALSE, 
  where = '!tbp'
  )
@

@@

Data retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}: \code{all\_params} returns the most recent records of all parameters at a station, \code{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \code{single\_param} is identical to \code{all\_params} except that a single parameter is requested. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.

@@

The second approach for data retrieval is to use the \code{import\_local} function to import data into R that are locally available after downloading from CDMO.  This approach is most appropriate for large, specific data requests. The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.  The \code{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \code{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

@@

\subsection{The \code{"swmpr"} object class}

All data retrieval functions return a \code{"swmpr"} object that includes relevant data and several attributes describing the dataset.  The data include a \code{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).  Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.  The following shows an example of the raw data imported using \code{all\_params}.

@@

The attributes for a \code{"swmpr"} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).  These act as metadata that are used internally by many of the package functions and are updated as the data are processed. The attributes are not visible with the raw data but can be viewed as follows.

@@

# view all attributes of dat
atts <- paste0('\\code{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\code{data.frame} object class.',
  'Row names of the data set, inherited from the \\code{data.frame} object class.',
  'Class of the data object indicating \\code{"swmpr"} and \\code{data.frame}.',
  'Station identifier used by NERRS as a string with 7 or 8 characters.',
  "Character vector of column names for data parameters, e.g., \\code{`do\\_mgl'}, \\code{`turb'}, etc.",
  'Indicates if QAQC columns are present in the raw data.',
  'Start and end dates for the data.',
'Timezone of the station using the city/country format\\textsuperscript{a}.',
  'Class of the \\code{datetimestamp} column, usually POSIXct unless data have been aggregated.'
)

to_tab <- data.frame(Attributes = atts, Class = atts_class, Description = atts_desc, stringsAsFactors = F)
foot <- c('\\textsuperscript{a}\\footnotesize Time zones that do not observe daylight savings are used for \\code{"swmpr"} objects and may not be cities in the United States.  For example, ``America/Jamaica" is used for Eastern Standard Time.')

@@

latex(
  to_tab[, -1, drop = F],
  file = '',
  caption = 'Attributes of a \\code{"swmpr"} object that describe characteristics of the data.',
  rowlabel = 'Attributes',
  colheads = c('Class', 'Description'),
  rowname = to_tab$Attributes,
  caption.loc = 'top',
  insert.bottom = foot,
  col.just=c("p{0.75in}", "p{3.25in}"), 
  label = 'tab:attributes', 
  table.env = FALSE, 
  where = '!tbp'
  )
@

The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.  A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.  Available methods for the \code{"swmpr"} class are described below and can also be viewed:

@@

A sample dataset can be downloaded for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.  These data are also included with the package as binary data files (RData) that can be loaded using the \code{data} function, as shown above.  These include \code{"swmpr"} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \code{?apacpnut}).

@@

The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{do\_mgl}, \code{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \code{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \code{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \code{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

funcs <- c('comb', 'qaqc', 'qaqcchk', 'rem\\_reps', 'setstep', 'subset')
funcs <- paste0('\\code{', funcs, '}')

descrips <- c(
  'Combines \\code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove QAQC columns and remove data based on QAQC flag values for a \\code{"swmpr"} object.  Only applies if QAQC columns are present. ', 
  'View a summary of the number of observations in a \\code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \\code{comb}.',
  'Subset by dates and/or columns for a \\code{"swmpr"} object.  This is a method passed to the generic \\code{subset} function provided in the base installation.'
  )

@@

Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \code{qaqc}.  The \code{qaqcchk} function can be used to view this information.

@@

funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'map\\_reserve', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
funcs <- paste0('\\code{', funcs, '}')
